---
title: "Quality control for PDAC single-cell transcriptomic data"
author: "Mohammadali (Sam) Khalilitousi, Yongjin P. Park"
date: "`r Sys.Date()`"
font_size: 12pt
---

```{r include = FALSE}
source("Util-rmd.R")
data.table::setDTthreads(15)
library(patchwork)
library(mmutilR)

fig.dir <- "Figures/QC"
.mkdir(fig.dir)
```

# Quality Control

## Data Sources

* FASTQ files for pancreatic ductal adenocarcinoma (PDAC) were obtained from the GSA: `https://ngdc.cncb.ac.cn/gsa/browse/CRA001160`

* FASTQ files were transformed into spliced and unspliced count matrices using kb-python: `https://www.kallistobus.tools/`

* The ENSEMBL Gene IDs were converted to Gene Symbols using biotools: `https://www.biotools.fr/human/ensembl_symbol_converter`

* Pancreatic Cancer associated genes: `https://maayanlab.cloud/Harmonizome/gene_set/pancreatic+cancer/DISEASES+Text-mining+Gene-Disease+Assocation+Evidence+Scores`

## Sample Annotation

Samples that contained different number of cell counts in their spliced and unsplcied gene count matrices were not selected for further analysis.

* Normal Samples: CRR034530, CRR034529, CRR034528, CRR034527,
  CRR034526, CRR034525, CRR034524, CRR034523, CRR034522, CRR034521,
  CRR034520

* Cancer Samples:

	* PDACI: CRR241799 (T3), CRR034500 (T5), CRR034512 (T17),
      CRR034513 (T18), CRR034519 (T24), CRR034517 (T22), CRR034516
      (T21), CRR241802 (T19), CRR034505 (T10)

	* PDACII: CRR241798 (T2), CRR034499 (T4), CRR034501 (T6),
      CRR034504 (T9), CRR241803 (T23), CRR241804 (T20), CRR241800
      (T7), CRR034511 (T16), CRR034506 (T11), CRR034510 (T15),
      CRR034509 (T14), CRR241801 (T13)

	* PDACIII: CRR241805 (T1), CRR034503 (T8), CRR034507 (T12)

* Samples we though had unknown stage: CRR034519 (T24), CRR241803
  (T23), CRR034517 (T22), CRR034516 (T21), CRR241804 (T20), CRR241800
  (T7), CRR241802 (T19), CRR034505 (T10), CRR034511 (T16), CRR034506
  (T11), CRR034510 (T15), CRR034509 (T14), CRR241801 (T13)

* Samples excluded: CRR034501 (T6), CRR034507 (T12), CRR034509 (T14),
  CRR034510 (T15), CRR034517 (T22)

## Data preparation

```{r echo=FALSE}
velocytoFolder <- "Data/Merged/"

.read.vec <- function(.file, .folder = velocytoFolder){
    fread(str_c(.folder, .file), header = FALSE) %>%
        unlist(use.names = FALSE)
}

barcodes <- .read.vec("samples.tsv")
features <- .read.vec("allCombinedgenes.tsv")
GWASgenes <- .read.vec("GWASgenes.txt") %>% intersect(features)

## We may not need to load entire MTX
## X <- Matrix::readMM(paste(velocytoFolder,"allCombined.mtx", sep=""))
## colnames(X) <- barcodes[,1]
## rownames(X) <- features[,1]
```

### Compute row-wise and column-wise scores of the combined single-cell data matrix


```{r}
.mkdir("Results/QC")
if(file.exists("Results/QC/MT.mtx.gz")) {
    .mt.only <- list(mtx = "Results/QC/MT.mtx.gz",
                     row = "Results/QC/MT.rows.gz",
                     col = "Results/QC/MT.cols.gz")
} else {
    .mt.genes <- features[str_starts(features, "MT-")]
    .mt.only <-
        rcpp_mmutil_copy_selected_rows("Data/Merged/allCombined.mtx.gz",
                                       str_c(velocytoFolder,"allCombinedgenes.tsv"),
                                       str_c(velocytoFolder,"samples.tsv"),
                                       .mt.genes,
                                       "Results/QC/MT")
}
```

```{r}
.mkdir("Table/QC/")
.file <- "Table/QC/cell_scores.csv.gz"
.file.2 <- "Table/QC/feature_scores.csv.gz"



if(!all(file.exists(c(.file, .file.2)))) {

    all.scores <-
        rcpp_mmutil_compute_scores("Data/Merged/allCombined.mtx.gz",
                                   str_c(velocytoFolder,"allCombinedgenes.tsv"),
                                   str_c(velocytoFolder,"samples.tsv"))

    mt.scores <-
        rcpp_mmutil_compute_scores(.mt.only$mtx,
                                   .mt.only$row,
                                   .mt.only$col)

    score.dt <-
        setDT(all.scores$col) %>%
        left_join(setDT(mt.scores$col),
                  by = "name",
                  suffix = c("", ".MT")) %>%
        as.data.table
    score.dt[is.na(`sum.MT`), `sum.MT` := 0]
    score.dt[, mito := `sum.MT` / `sum` * 100]
    fwrite(score.dt, .file)

    row.score.dt <- setDT(all.scores$row)
    fwrite(row.score.dt, .file.2)

} else {
    score.dt <- fread(.file)
    row.score.dt <- fread(.file.2)
}
```

```{r}
MITO.CUTOFF <- 50
NNZ.CUTOFF <- 300
```

#### Select high-quality cells that contain at least $>$ `r NNZ.CUTOFF` non-zero genes expressed with the mitochondrial gene activity $<$ `r MITO.CUTOFF` %

```{r}
.hist.nnz <- score.dt[, .(.N), by = .(0.1 * ceiling(10 * log10(nnz)))]
names(.hist.nnz) <- c("nnz.log10", "count")
.hist.mito <- score.dt[, .(.N), by = .(ceiling(mito/2) * 2)]
names(.hist.mito) <- c("mito", "count")

.dt <- score.dt %>%
    mutate(nnz.log10.grid = 0.1 * ceiling(10 * log10(nnz)),
           mito.grid = ceiling(mito))

.hist.joint <- .dt[, .(.N), by = .(nnz.log10.grid, mito.grid)]
```

```{r echo = FALSE, results = "asis", fig.width = 5, fig.height=4}
.lab <- function(x) num.sci(10^(x))

p1 <-
    .gg.plot(.hist.nnz, aes(x = `nnz.log10`, y = `count`)) +
    geom_bar(stat="identity") +
    geom_vline(xintercept = log10(NNZ.CUTOFF), colour = "red", lty = 2, size = .5) +
    scale_x_continuous("# non-zero elements per cell", labels = .lab) +
    scale_y_continuous(labels = num.int)

p2 <-
    .gg.plot(.hist.mito, aes(y = `mito`, x = -`count`)) +
    geom_segment(aes(yend = `mito`, xend = 0), size = 1, colour = "gray30") +
    geom_hline(yintercept = MITO.CUTOFF, colour = "red", lty = 2, size = .5) +
    theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +
    ylab("Mitochondrial activity (%)") +
    scale_x_continuous("count", labels = function(x) num.sci(-x))

p3 <-
    .gg.plot(.hist.joint, aes(x = nnz.log10.grid, y = mito.grid, fill = log10(`N`))) +
    theme(panel.background = element_blank()) +
    scale_x_continuous("# non-zero elements per cell", labels = .lab) +
    scale_y_continuous("Mitochondrial activity (%)") +
    geom_tile() +
    geom_vline(xintercept = log10(NNZ.CUTOFF), colour = "red", lty = 2, size = .5) +
    geom_hline(yintercept = MITO.CUTOFF, colour = "red", lty = 2, size = .5) +
    scale_fill_distiller("count", direction = 1, labels = function(x) num.int(10^x)) +
    theme(legend.position = c(1,1), legend.justification = c(1,1))

p0 <- ggplot() + theme_void()

ptop <- (p0|p1) + plot_layout(widths=c(1,3))
pbot <- (p2|p3) + plot_layout(widths=c(1,3))
plt <- (ptop/pbot) +plot_layout(heights=c(1,3))
print(plt)

.file <- str_c(fig.dir, "/Fig_qc_nnz_mito_cutoff.pdf")
.gg.save(.file, plot = plt, width = 5, height = 4)
```

```{r include = FALSE}
ntot.cells <- nrow(score.dt)
nqc.cells <- nrow(score.dt[mito < MITO.CUTOFF & nnz > NNZ.CUTOFF])
```

#### Of `r num.int(ntot.cells)` cells, we retained `r num.int(nqc.cells)` (`r num.round(nqc.cells/ntot.cells*100)` %), discarding `r num.int(ntot.cells - nqc.cells)` cells (`r num.round(100 * (ntot.cells - nqc.cells) / ntot.cells)` %) by applying the Q/C steps

### We focused on the known PDAC-associated genes implicated by the previous genome-wide association studies

```{r}
GWASgenes <- .read.vec("GWASgenes.txt")
```

