---
title: "Topic modelling for PDAC single-cell transcriptomic data"
author: "Mohammadali (Sam) Khalilitousi, Yongjin P. Park"
date: "`r Sys.Date()`"
font_size: 12pt
---

```{r include = FALSE}
#source("Util-rmd.R")
#data.table::setDTthreads(15)
#library(patchwork)
#library(mmutilR)

fig.dir <- "Figures/QC"
#.mkdir(fig.dir)
velocytoFolder <- "Data/Merged/"
```

# Topic Modeling

## Data Sources

* QC PDAC_GWAS and QC PDAC_GWAS_BG count matrices obtained from the previous step (STEP1_QC_DATA.rmd): `https://causalpathlab.github.io/pdac_velocity_topic/STEP1_QC_Data.html`
* Summary GWAS of invasive pancreatic ductal adenocarcinoma from the JaPAN consortium: `http://www.aichi-med-u.ac.jp/JaPAN/current_initiatives-e.html`

## Data Preparation

## Put the new data preparation here (don't load everything)


## Running Topic Modeling Code
```{r}
llibrary(Matrix)
library(fastTopics)

X <- Matrix::readMM("/projects/molonc/park_lab/sam/pancreasTopic/scripts/QC/yppQC/QC/PDAC_GWAS.mtx.gz")
barcodes <- read.table("/projects/molonc/park_lab/sam/pancreasTopic/scripts/QC/yppQC/QC/PDAC_GWAS.cols.gz")
features <- read.table("/projects/molonc/park_lab/sam/pancreasTopic/scripts/QC/yppQC/QC/PDAC_GWAS.rows.gz")
colnames(X) <- barcodes[,1]
rownames(X) <- features[,1]

ngenes = as.numeric(dim(X)[1])

#separating spliced and unspliced counts
s = X[1:(ngenes/2),]
u = X[(ngenes/2+1):ngenes,]

#Finding ratios between spliced and unspliced counts and vice versa
ratio1 = log(u+1)-log(s+1)
rownames(ratio1) = paste(gsub('.{2}$', '', rownames(s)),"u/s",sep = "_")
ratio1[ratio1<0] = 0 #get rid of negative values
ratio2 = log(s+1)-log(u+1)
rownames(ratio2) = paste(gsub('.{2}$', '', rownames(s)),"s/u",sep = "_")
ratio2[ratio2<0] = 0 #get rid of negative values
X = rbind(ratio1,ratio2)

#removing rows with sum = 0
X = X[rowSums(X)>0,]
#removing colums with sum = 0
X = X[,colSums(X)>0]

#Putting th ratios matrix in the cells x genes format preferred by fastTopics
X=t(X)
X = as(X, "dgCMatrix")

fit10 <- fastTopics::fit_topic_model(X, k = 10, numiter.main = 200, numiter.refine = 200)
save(fit10, ascii=FALSE, file="fit10_GWAS")

fit10DE = fastTopics::diff_count_analysis(fit10,X)
save(fit10DE, ascii=FALSE, file="fit10DE_GWAS")
```



## Analyzing Topic Modeling Results
```{r}
set.seed(1)

#These topics were found using the s/u & u/s ratios.
load("/Users/samtoussi/Desktop/yppQC/fit10_GWAS")
#load("/Users/samtoussi/Desktop/Pancreas/LDAimplementation/upload/fit10")

topic_colors10 <- c("gold","forestgreen","dodgerblue","green","skyblue", "pink", "red", "brown", "purple","orange" )
structure_plot(fit10,colors = topic_colors10) #view structure plot

#Labeling cells by their cancer stage
cellNames =rownames(fit10[["L"]])
caseControl = as.data.frame(cellNames)
caseControl$labels = "Control"
rownames(caseControl)= caseControl$cellNames
PDACI = 'CRR241799|CRR034500|CRR034512|CRR034513|CRR034519|CRR034517|CRR034516|CRR241802|CRR034505'
PDACII = 'CRR241798|CRR034499|CRR034501|CRR034504|CRR241803|CRR241804|CRR241800|CRR034511|CRR034506|CRR034510|CRR034509|CRR241801'
PDACIII = 'CRR241805|CRR034503|CRR034507'
caseControl[grep(PDACI,rownames(caseControl),value=TRUE),"labels"] = "PDACI"
caseControl[grep(PDACII,rownames(caseControl),value=TRUE),"labels"] = "PDACII"
caseControl[grep(PDACIII,rownames(caseControl),value=TRUE),"labels"] = "PDACIII"

#view structure plot, grouped by case/control label
structure_plot(fit10,colors = topic_colors10,grouping = caseControl$labels, gap = 25) #view structure plot


#Cluster cells based on topics
set.seed(1)
pca10 <- prcomp(fit10$L)$x
clusters10GWAS <- kmeans(pca10,centers = 15,iter.max = 10000,)$cluster
structure_plot(fit10,topics = 1:10,colors = topic_colors10,gap = 25, grouping = clusters10GWAS)
```

## Running Topic Modeling on Entire Canonical Pathways instead of Individual Genes 
```{r}
library(msigdbr) #loading the Molecular Signatures Database package

test  = msigdbr_collections()

X <- Matrix::readMM("QCcounts.mtx")
barcodes <- read.table("QCcells.txt")
features <- read.table("QCgenes.txt")
colnames(X) <- barcodes[,1]
rownames(X) <- features[,1] 

#s = X[1:58367,]
#u = X[58368:116734,]

s = X[1:1434,]
u = X[1435:2868,]

#obtaining and combining all the CP pathways
canonicalPatways = rbind(msigdbr(species = "Homo sapiens",category = "C2",subcategory = "CP:BIOCARTA"),
                         msigdbr(species = "Homo sapiens",category = "C2",subcategory = "CP:KEGG"),
                         msigdbr(species = "Homo sapiens",category = "C2",subcategory = "CP:REACTOME"))

canonicalPatways = canonicalPatways[,c(3,4)] #removing unnecessary data columns

#creating 2 versions: one ending with _s for spliced genes and one ending with _u for unspliced	genes
CPs=canonicalPatways
CPu=canonicalPatways
CPs$gene_symbol = paste(CPs$gene_symbol,"s",sep="_")
CPu$gene_symbol = paste(CPu$gene_symbol,"u",sep="_") 

combined = rbind(CPs,CPu)

#in order to get a matrix consisting of pathways and their corresponding array of genes (nPathways x 2 matrix)
pathwayCounts = aggregate(combined$gene_symbol, by=list(Category=combined$gs_name), FUN=paste)

geneListS=rownames(s)
geneListU=rownames(u)
#Creating pathway x genes (s and u separately) matrix
i = vector()
j = vector()
for (x in 1:length(pathwayCounts$x)) {
j = c(j, which(geneListS %in% unlist(pathwayCounts$x[ x ])))
i = c(i,rep(x,each=length(which(geneListS %in% unlist(pathwayCounts$x[ x ])))))
}
sp_matrixS <- sparseMatrix(i=i,j=j,x=1,dims=list(length(pathwayCounts$x),length(geneListS)))
Matrix::writeMM(sp_matrixS,"sp_matrixS")

i = vector()
j = vector()
for (x in 1:length(pathwayCounts$x)) {
j = c(j, which(geneListU %in% unlist(pathwayCounts$x[ x ])))
i = c(i,rep(x,each=length(which(geneListU %in% unlist(pathwayCounts$x[ x ])))))
}
sp_matrixU <- sparseMatrix(i=i,j=j,x=1,dims=list(length(pathwayCounts$x),length(geneListU)))
Matrix::writeMM(sp_matrixU,"sp_matrixU")

#Obtaining the pathway x cells matrices
pathwayXs = sp_matrixS %*% s
pathwayXu = sp_matrixU %*% u

rownames(pathwayXs) = paste(pathwayCounts[,1],"s", sep = "_")
rownames(pathwayXu) = paste(pathwayCounts[,1],"u", sep = "_")
colnames(pathwayXs) = colnames(X)
colnames(pathwayXs) = colnames(X)

pathwayXs = as(pathwayXs, "dgCMatrix")
Matrix::writeMM(pathwayXs, "pathwayXs.mtx")

pathwayXu = as(pathwayXu, "dgCMatrix")
Matrix::writeMM(pathwayXu, "pathwayXu.mtx")

write.table(rownames(pathwayXs), "pathwayNamesS.txt", quote = F,row.names = F,col.names = F)
write.table(rownames(pathwayXu), "pathwayNamesU.txt", quote = F,row.names = F,col.names = F)

#running topic modelling just as before ----------------------------------------------------

ratio1 = log(pathwayXu+1)-log(pathwayXs+1)
rownames(ratio1) = paste(gsub('.{2}$', '', pathwayCountsS[,1]),"u/s",sep = "_")
ratio1[ratio1<0] = 0 #get rid of negative values

ratio2 = log(pathwayXs+1)-log(pathwayXu+1)
rownames(ratio2) = paste(gsub('.{2}$', '', pathwayCountsS[,1]),"s/u",sep = "_")
ratio2[ratio2<0] = 0 #get rid of negative values

X = rbind(ratio1,ratio2)

#removing rows with sum = 0
X = X[rowSums(X)>0,]
#removing colums with sum = 0
X = X[,colSums(X)>0]

X = t(X) #transpose
X = as(X, "dgCMatrix")

pathwayTopic <- fastTopics::fit_topic_model(X,k = 10)
save(pathwayTopic,ascii = F,file = "pathwayTopic")

pathwayTopicDE = fastTopics::diff_count_analysis(pathwayTopic,X)
save(pathwayTopicDE, ascii=FALSE, file="pathwayTopicDE")

```


## Entropy calculation
```{r}
#Formula for entropy of X: H(X) = - sum(P(xi)*log(P(xi)))

Entropy = apply(fit10$L, 1, function(x) - sum(x*log(x)))
hist(Entropy,breaks = 100)

Entropy = apply(combinedFit$L, 1, function(x) -sum(x*log(x)))
hist(Entropy,breaks = 100)

Entropy = apply(combinedFit2$L, 1, function(x) -sum(x*log(x)))
hist(Entropy,breaks = 100)

#CALCULATE STATISTICS

```

## Validation on the Japanese dataset
```{r}
#as seen from the structure plots with case and control labels, some topics are more present in cells which belong to the control samples and vice versa.
#Control group: 3,7,6
#Disease group: 10,1,2
library(htmltools)
library(DESeq2)
library(ggplot2)

X <- Matrix::readMM("/Users/samtoussi/Desktop/yppQC/QC/PDAC_GWAS_BG.mtx.gz")
barcodes <- read.table("/Users/samtoussi/Desktop/yppQC/QC/PDAC_GWAS_BG.cols.gz")
features <- read.table("/Users/samtoussi/Desktop/yppQC/QC/PDAC_GWAS_BG.rows.gz")
colnames(X) <- barcodes[,1]
rownames(X) <- features[,1]

dds  = DESeqDataSetFromMatrix(countData=X, colData =caseControl , design =~labels)
dds = DESeq(dds)

res = results(dds)

#import results instead of running pipeline
res = read.table("/Users/samtoussi/Desktop/yppQC/manualDEresults.txt")


#Analyzing results
head(results(dds, tidy = T))
head(res[order(res$pvalue),])#sorting by p-value


#vieweing results
par(mfrow = c(2,3))
plotCounts(dds, gene="AK2_s", intgroup="V1")


par(mfrow=c(1,1))
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot", xlim=c(-3,3)))
with(subset(res, padj<.01 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(res, padj<.01 & abs(log2FoldChange)>0.5), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))




```


